Competing Constructions for German Measure Noun Phrases: from Usage Data to Experimental Validation

### 2017-04-03 Latex/article.tex
evening sync

This paper deals with a morpho-syntactic alternation that occurs only in a very specific syntactic measure noun phrase construction.
By \textit{alternation} I refer to a situation where two or more forms or constructions are available with no clear difference in acceptability, function, or meaning.

In German grammar, the term \textit{Zweifelsfälle} `cases of doubt' is often used \citep{Klein2009}.

@@

%%%%%% From old structure:
%
%\subsection{Cognitively oriented corpus linguistics}
%\label{sec:cocl}
%
% Mention: Corpus-as-input is OK as a hypothesis but NOT as a research programme
% Entrenchment_Stefanowitsch and Flach for submission + Schmid C2C
%  => otherwise any spurious pattern found in corpora (meeting certain criteria)
%  => is automatically "cognitive"
%
%Following the descriptive introduction to the measure noun phrase alternation in Section~\ref{sec:germanmeasurenps}, this section situates the work presented here within the broader theoretical context of what can be called \textit{cognitively oriented corpus linguistics}.
%
%Unless there are strong, numeric, and cognitively motivated hypotheses, some experimental validation is required to make a corpus study \textit{cognitive} in any sense.
%Assuming the \textit{corpus as input} hypothesis without experimental validation is completely unsound in as much as researcher could extract any arbitrary -- often spurious -- usage pattern from a (usually noisy) corpus and claim that it has a cognitive reality.
%
% Cite:
% [Cognitive Linguistics] Similarity in linguistic categorization- The importance of necessary properties
% 10.1.1.980.9935
% __IMP__Measurenouns__cog-2015-0101
%
% Maybe:
% [Cognitive Linguistics] The processing of verb-argument constructions is sensitive to form, function, frequency, contingency and prototypicality
% [Cognitive Linguistics] Alternation-based generalizations are stored in the mental grammar- Evidence from a sorting task experiment //// along with CApelle's allostructions
%
% Method:
% Noble

@@

In this section, I briefly motivate the choice of statistical models which I use later in Section~\ref{sec:corpusstudies}.

% NO to exploratory methods if we have hypotheses
% Testing (hypotheses about signs of regressors)
% Estimation => bootstrapping
% Model selection
%  => random vs fixed is not so simple, rather matter of getting model to converge
%  => Maximality
%  => NO to Gries 2015
% Model evaluation
% VS: Bayesian hype (Levshina)
% VS: alternative math (R-W)

% Cite:
% __IMP__Measurenouns__cog-2015-0054 (Levshina)
% [Cognitive Linguistics] Towards cognitively plausible data science in language research (NDL)

% Stats:
% BarrEa_RandomEffectsStructure_2014_JMemLang
% ContentServer (1)
% importatn_quotes
% Menard_2000-Coefficients_of_Determination_for_Multiple_Logistic_Regression_Analysis

%In the following, I examine the case alternation in German measure NPs under a cognitively oriented corpus linguistic perspective.
%Since, however, the statistical tool kit (hierarchical generalised linear models, general tools such as $p$ values and confidence intervals, resampling techniques, model selection and model evaluation strategies, alternative algorithms for estimation, etc.) is heterogeneous and often criticised (especially under noisy conditions and when researcher run the risk of doing data mining instead of proper theory testing) I present a careful protocol for specifying, estimating, validating, and interpreting such models in Section \ref{sec:corpusstudies}.

Before discussing the results in Section~\ref{sec:corpusstudies}, I show how I made sure that the relevant data is contained in the sample in Section \ref{sec:gettingdata}.

%\subsection{Tools for modelling alternations}
%\label{sec:righttools}
%
%In this section, I discuss the relevant statistical approaches to modelling alternations based on corpus data.
%I mainly focus on well-established Generalised Linear Models (GLMs) and their \textit{multi-level} or \textit{hierarchical} extensions, which are also called \textit{mixed models} (GLMMs).
%
%% TODO CONTENT HERE
%
%In Section~\ref{sec:corpushierarchicalmodel}, I present the hierarchical model which is actually interpreted as the primary result of this study.
%Section~\ref{sec:bayesian} takes a short detour and shows (by example) that results from Bayesian estimation are not intrinsically different from those obtained from Maximum Likelihood methods, although Bayesian approaches have recently been touted to solve various problems facing applied statistics in various fields, even including cognitively oriented corpus linguistics.
%Finally, I summarise and interpret the results of the corpus study in Section~\ref{sec:interpretation}.

% VanderPlas: Bayesian

@@

In a probabilistic framework like Prototype Theory, the crucial question is, however, what controls speakers' decisions to use either variant beyond simple lemma preference effects. 
I argue that the NAC and the PGC are prototypes associated with different degrees of the grammaticalisation of the measure noun, morpho-syntactic properties, and register effects.

@@

Furthermore, the grammaticalisation path as described above leads from NPs denoting individuated objects standing in a \textit{part-of} relation to a construction with a more diffuse \textit{quantity-of} relation.
Both types of relations can be numerically quantified -- in as much as a precise number of \textit{parts} or a numerically exact \textit{quantity} can be specified --, but it is much more prototypical of quantities to be specified with numerical precision.
Since the \NACa\ is more closely associated with the \textit{quantity-of} relation, cardinals as attributes of the measure noun are expected to have a higher proportional frequency in the \NACa.
For illustration, (\ref{ex:determiners}) shows the expected variants under this hypothesis.
In (\ref{ex:determiners:nac}), the measure noun is modified by a cardinal \textit{drei} `three', and hence the \NACa\ (the kind NP shows case agreement) is preferred.
In (\ref{ex:determiners:pgc}), the measure noun is modified by a non-cardinal determiner \textit{einige} `some', and the \PGCa\ (the kind NP is in the genitive) is preferred.

@@

Finally, it has often been iterated that the \PGCa\ is more typical of higher registers or even exclusive to written language (see \citealp[320--323]{Hentschel1993}).
This is not surprising in as much as the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial vernacular variants of German as a result of a diachronic process wherein many (but not all) uses of the genitive are replaced by other cases or periphrastic constructions \citep{FleischerSchallert2011}.
Under an integral view of prototypes, which incorporate effects related to larger contexts and registers, such preferences can be to be part of the construction prototypes.
In the corpus study in Section~\ref{sec:corpusstudies}, register effects will therefore be modelled -- even if only using some very simple proxies.

@@

For the present study, I used the German \textit{Corpus from the Web} (COW) in its 2014 version DECOW14A (\citealp{SchaeferBildhauer2012full,Schaefer2015b} and \citealp{BiemannEa2013,SchaeferBildhauer2013} for overviews of web corpora in general and the methodology of their construction), which contains almost 21 billion tokens.%
\footnote{The corpora are made available for free at \url{https://www.webcorpora.org}.
At the time of this writing, a newer 2016 version DECOW16 had already been released.}
I chose this corpus for two main reasons.
First, the external validity of any study is increased through a higher heterogeneity of the sample \citep[30]{MaxwellDelaney2004}, and the DECOW corpus has clearly a much more heterogeneous composition compared to the only other very large corpus of German, the DeReKo \citep{KupietzEa2010} of the Institute for the German Language (IDS), which contains almost exclusively newspaper texts.%
\footnote{It was shown in \cite{W16-2601} that, for example, the spread of topics is much smaller in DeReKo compared to DECOW.}
Second, it was already mentioned that normative grammars often adopt clear positions regarding the grammaticality of either the \NACa\ or the \PGCa.
Thus, newspaper text or any other text that conforms strongly to normative grammars might not represent the alternation phenomenon fully (and without bias) because authors and proof-readers might favour one alternative or the other.
Web corpora, on the other hand, contain at least some amount of non-standard language from forums and similar sources.
For these or similar reasons, COW corpora have been used in a number of peer-reviewed publications, for example \cite{VanGoethemHiligsmann2014,VanGoethemHuening2015,MuellerS2014,Schaefer2016c,SchaeferSayatz2014,SchaeferSayatz2016,Zimmer2015}. 
Therefore, DECOW can be considered the obvious choice for this study.%
\footnote{The use of web data for linguistic research does require explicit and careful justification.
Due to the noisy nature and unknown composition of the web, only carefully designed and established web corpora like the COW corpora or the SketchEngine corpora \citep{KilgarriffEa2014} should be used.
Clearly, using search engine results is \textit{bad science} for many reasons, most prominently total irreproducibility of results, as \cite{Kilgarriff2006} pointed out more than ten years ago.
Careless use of search engine results is still found, however, see \citet[171--175]{DeclerckBrems2016}.}

@@

The coefficient estimates are specified in Table~\ref{tab:bigtable} for each regressor (or regressor level) in the columns labelled \textit{Coefficient}.
For a robust quantification of the precision of the estimation, I ran a parametric bootstrap (using the \textit{confint.merMod} function from \textit{lme4}) with $1,000$ replications, and using the percentile method for the calculation of the intervals.
The resulting 95\% bootstrap confidence intervals are reported in Table~\ref{tab:bigtable} in the columns labelled \textit{CI low} and \textit{CI high} (= upper and lower 2.5th percentiles).
The column \textit{CI contains 0} shows an asterisk for those intervals that do \textit{not} include 0.
Furthermore, for each regressor, a $p$ value was obtained by dropping the regressor from the full model, re-estimating the nested model and comparing it to the full model.
Instead of inexact Wald approximations or Likelihood Ratio Tests, I used a drop-in bootstrap replacement for the Likelihood Ratio Test from the function \textit{PBmodcomp} from the \textit{pbkrtest} package \citep{HalekohHojsgaard2014}, hence I call the value $p_{\text{PB}}$.
These bootstrapped $p$ values are given in the columns labelled p\Sub{PB} in Table~\ref{tab:bigtable}.
Only \textit{Kindfreq} ($\mpPB=0.095$) can be seen as slightly too high to be convincing (`non-significant').

>>>>

This paper deals with a morpho-syntactic alternation that occurs only in a very specific syntactic measure noun phrase construction in German.
By \textit{alternation} I refer to a situation where two or more forms or constructions are available with no clear difference in acceptability, function, or meaning.

% In German grammar, the term \textit{Zweifelsfälle} `cases of doubt' is often used \citep{Klein2009}.

@@

%%%%%% From old structure:
%
%\subsection{Cognitively oriented corpus linguistics}
%\label{sec:cocl}
%
% Mention: Corpus-as-input is OK as a hypothesis but NOT as a research programme
% Entrenchment_Stefanowitsch and Flach for submission + Schmid C2C
%  => otherwise any spurious pattern found in corpora (meeting certain criteria)
%  => is automatically "cognitive"
%
% Cite:
% [Cognitive Linguistics] Similarity in linguistic categorization- The importance of necessary properties
% 10.1.1.980.9935
% __IMP__Measurenouns__cog-2015-0101
%
% Maybe:
% [Cognitive Linguistics] The processing of verb-argument constructions is sensitive to form, function, frequency, contingency and prototypicality
% [Cognitive Linguistics] Alternation-based generalizations are stored in the mental grammar- Evidence from a sorting task experiment //// along with CApelle's allostructions
%
% Method:
% Noble

@@

In this section, I briefly motivate the choice of statistical models which I use later in Section~\ref{sec:corpusstudies}, namely Generalised Linear Mixed Models (GLMMs) using Maximum Likelihood Estimation.%
\footnote{It would be technically more precise to speak of Hierarchical Logistic Regression rather than a GLMM in the case of this study.}
Many readers might think that the use of this class of statistical methods in the modeling of grammatical alternations does not require much motivation.
After all, GLMMs have been established as the major tool in the analysis of (\citealp{BresnanEa2007,BresnanHay2010,BresnanFord2010,DivjakArppe2013,Gries2015,NessetJanda2010}, to name only a few publications).
However, over the past years alternative methods have been proposed.
Within the Generalised Linear Modeling framework, multimodel averaging and multimodel inference (\citealp{BurnhamAnderson2002}, linguistic applications include \citealp{BarthKapatsinski2014,KupermanBresnan2012}) have been suggested to deal with model uncertainty and the sometimes problematic necessity to settle on one specific model structure.
Also, alternative estimators have been introduced into the linguistics literature, most notably Bayesian estimation (a comprehensive introduction in \citealp{GelmanEa2014}, see \citealp{Levshina2016,Divjak2016} for praises by linguists).
On the other hand, completely different modeling approaches have been proposed, for example Trees \& Forests \citep{StroblEa2009} and Naive Discriminative Learning \citep{Baayen2011,MilinEa2016}.
The last two methods are summarised and compared to regression methods in \cite{BaayenEa2013}, and \cite{TheijssenEa2013} provides a similar comparison with even more methods.

Often, linguists provide convincing arguments why adopting a statistical tool might beneficial.
For example, \citep{Baayen2011,MilinEa2016} argue that Naive Discriminative Learning (NDL) is more adequate for modeling in strands of linguistics where cognitive reality is desired.
For example, the fact that regression models cannot properly deal with redundancy in the form of multicollinearity makes them implausible given some cognitive research, while NDL can deal with redundant information \citep[508]{MilinEa2016}.
Furthermore, the fact that likelihood maximisation is a form of optimsation is criticised \citep[508]{MilinEa2016}, although in NDL and Trees \& Forests are praised for \textit{arriv}[\textit{ing}] \textit{at their optimal solutions on their own} \cite[256]{BaayenEa2013}.
More fundamentally, regression modeling has been criticised because it usually relies on high-level linguistic abstractions, and NDL (or other methods) can achieve at least the same adequacy by considering low-level units such as words or even character ngrams <++>

@@

In a probabilistic framework like Prototype Theory, the crucial question (beyond simple lemma preference effects) is, however, what controls speakers' decisions to use either variant. 
In the remainder of this section, I argue that the NAC and the PGC are prototypes associated with different degrees of the grammaticalisation of the measure noun, a related morpho-syntactic property, and register effects.

@@

Furthermore, the grammaticalisation path as described above leads from NPs denoting individuated objects standing in a \textit{part-of} relation to a construction with a more diffuse \textit{quantity-of} relation.
Both types of relations can be numerically quantified -- in as much as a precise number of \textit{parts} or a numerically exact \textit{quantity} can be specified --, but it is much more prototypical of quantities to be specified with numerical precision.
Since the \NACa\ is more closely associated with the \textit{quantity-of} relation, cardinals as attributes of the measure noun are expected to have a higher proportional frequency in the \NACa.
For illustration, (\ref{ex:determiners}) shows the expected variants under this hypothesis.
In (\ref{ex:determiners:nac}), the measure noun is modified by a cardinal \textit{drei} `three', and hence the \NACa\ is preferred.
In (\ref{ex:determiners:pgc}), the measure noun is modified by a non-cardinal determiner \textit{einige} `some', and the \PGCa\ is preferred.

@@

Finally, it has often been iterated that the \PGCa\ is more typical of higher registers or even exclusive to written language (see \citealp[320--323]{Hentschel1993}).
This is not surprising in as much as the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial vernacular variants of German as a result of a diachronic process wherein many (but not all) uses of the genitive are replaced by other cases or periphrastic constructions \citep{FleischerSchallert2011}.
Under an integral view of prototypes, which incorporate effects related to larger contexts and registers, such preferences can be to be part of the construction prototypes.
In the corpus study in Section~\ref{sec:corpusstudies}, register effects will therefore be modelled -- even if only using two very simple proxy variables.

@@

For the present study, I used the German \textit{Corpus from the Web} (COW) in its 2014 version DECOW14A (\citealp{SchaeferBildhauer2012full,Schaefer2015b} and \citealp{BiemannEa2013,SchaeferBildhauer2013} for overviews of web corpora in general and the methodology of their construction), which contains almost 21 billion tokens.%
\footnote{The corpora are made available for free at \url{https://www.webcorpora.org}.
At the time of this writing, a newer 2016 version DECOW16 had already been released.}
I chose this corpus for two main reasons.%
\footnote{The use of web data for linguistic research does require explicit and careful justification.
Due to the noisy nature and unknown composition of the web, only carefully designed and established web corpora like the COW corpora or the SketchEngine corpora \citep{KilgarriffEa2014} should be used.
Clearly, using search engine results is \textit{bad science} for many reasons, most prominently total irreproducibility of results, as \cite{Kilgarriff2006} pointed out more than ten years ago.
Careless use of search engine results is still found, however, see \citet[171--175]{DeclerckBrems2016}.}
First, the external validity of any study is increased through a higher heterogeneity of the sample \citep[30]{MaxwellDelaney2004}, and the DECOW corpus has clearly a much more heterogeneous composition compared to the only other very large corpus of German, the DeReKo \citep{KupietzEa2010} of the Institute for the German Language (IDS), which contains almost exclusively newspaper texts.%
\footnote{It was shown in \cite{W16-2601} that, for example, the spread of topics is much smaller in DeReKo compared to DECOW.}
Second, it was already mentioned that normative grammars often adopt clear positions regarding the grammaticality of either the \NACa\ or the \PGCa.
Thus, newspaper text or any other text that conforms strongly to normative grammars might not represent the alternation phenomenon fully (and without bias) because authors and proof-readers might favour one alternative or the other.
Web corpora, on the other hand, contain at least some amount of non-standard language from forums and similar sources.
For these or similar reasons, COW corpora have been used in a number of peer-reviewed publications, for example \cite{VanGoethemHiligsmann2014,VanGoethemHuening2015,MuellerS2014,Schaefer2016c,SchaeferSayatz2014,SchaeferSayatz2016,Zimmer2015}. 
Therefore, DECOW can be considered the obvious choice for this study.

@@

The coefficient estimates are specified in Table~\ref{tab:bigtable} for each regressor (or regressor level) in the columns labelled \textit{Coefficient}.
For a robust quantification of the precision of the estimation, I ran a parametric bootstrap (using the \textit{confint.merMod} function from \textit{lme4}) with $1,000$ replications, and using the percentile method for the calculation of the intervals.
The resulting 95\% bootstrap confidence intervals are reported in Table~\ref{tab:bigtable} in the columns labelled \textit{CI low} and \textit{CI high} (= upper and lower 2.5th percentiles).
The column \textit{CI contains 0} shows an asterisk for those intervals that do \textit{not} include 0.
Furthermore, for each regressor, a $p$ value was obtained by dropping the regressor from the full model, re-estimating the nested model and comparing it to the full model.
Instead of inexact Wald approximations or Likelihood Ratio Tests, I used a drop-in bootstrap replacement for the Likelihood Ratio Test from the function \textit{PBmodcomp} from the \textit{pbkrtest} package \citep{HalekohHojsgaard2014}, hence I call the corresponding value $p_{\text{PB}}$.
These bootstrapped $p$ values are given in the columns labelled p\Sub{PB} in Table~\ref{tab:bigtable}.
Only \textit{Kindfreq} ($\mpPB=0.095$) can be seen as slightly too high to be convincing (`non-significant').
